Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-05-21 paper/paper.tex
Fix typos and mistakes

To run the program, we first use the @inject@ function (below) to construct an initial machine
state given an expression @e@. The initial state contains an empty environment,
store, and stack.
\begin{lstlisting}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{lstlisting}

The @drive@ function is then used to evaluate
to a final state by iteratively applying @step@ on the current state until a state is reached
ion which the control is an atomic expression and the continuation stack is empty.
Naturally, we can then extract the value from the final state at last.

@@

In the definition of @State@, the continuation store disappears, and
component @konts@ becomes a list of frames. The other components remain unchanged.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentions that for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
This poses a problem, as the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (i.e., the @Frame@ in our program) is \emph{not}
nondeterministic, and the worklist @todo@ actually implicitly handles the nondeterminism.
Thus, if we simply refunctionalize the frames to functions, it does not help us move toward
abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all nondeterministic choices.
This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second as a \emph{nondeterministic
continuation}.

@@

Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bind to the function's formal argument,
and the store, time, and frames at the fork point.
In the definition of @NDState@, an auxiliary function @toState@ is added that
converts itself to @State@.

@@

However, since we have added nondeterminism into state, an empty list of frames means
we have the reached the halt of \textit{one computation path}, and we should
determine the next state indicated by @NDCont@.
Thus, we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we have tried all the closures of this fork point and should move to the
next fork point. Since the @NDCont@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The nondeterministic continuation @new_ndk@ is also updated by the removal of that closure.
\end{itemize}

@@

With this, have a single function to perform both abstract evaluation and the collection of
intermediate states. Given @inject(e)@ as an initial state and an empty set as the
initial set of states reached, we can easily define the entrance function @analyze@ as
follows:

@@

@continue@ is mainly what we have seen for the second case in
@drive_step@, except that @ndcontinue@ is called when
the normal continuation is empty.
Since we are analyzing a language in ANF, the first-order data representation of
continuations is simplified to a Scala @List@.
An empty list @Nil@ represents the halt of a computation path, whereas
a cons @::@ means the top frame of the list holds the binding variable and
evaluation context.
An analogy to this in standard CEK machines would be an abstract class (e.g., @cont@) with
three variants: @halt@, @arg(value, cont)@, and @arg(exp, env, cont)@.

@@

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ implements the nondeterministic
evaluation, given the defined @nd@, the evaluation function @aeval@ can be easily
defined. In the case that @e@ is a @let@ expression,
we invoke @nd@ for twice and perform a nesting two-step, depth-first evaluation
over possible values of @App(f, ae)@.
The values of branches from on fork point will be accumulated, and then
returned to the upper fork point.

@@

Based on the refunctionalized AAM from the last section, we use the same nondeterministic
operator @nd@, but further extend it to a fixed-point cached evaluation in this section.
The fixed-point caching algorithm performs a sound over-approximation of all possible
concrete reachable paths and values, and also prevents non-termination when analyzing diverged programs.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantics artifact, then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order non-deterministic evaluation through higher-order functions representing continuations, 
but also drive us to use a new caching algorithm.

@@

\textbf{Nondeterminism Operators, Again.}
The first operator @nd@ is simplified from the previous one, but specialized with type @Ans@ and @Cache@. 
We may still keep the types to be generic, but the purpose of presentation clearly, exposing @Cache@ explictly
reveals that cache should be monotonically accumulated to the rest of computation.
Now @nd@ only takes a function @f@ that applies on type @(T, Cache)@ where value of type @T@ comes from the set,
and @nd@ does recursive call on itself over elements in the set @ts@.
We also move the accumulation operation (@++@) inside of @nd@.

@@

The function @ndcps@ takes two arguments: a set of elements with type @T@, 
and an initial accumulated value of type @Ans@.
The @ndcps@ iteratively returns an element from the set along with the latest cache 
, and the return type is also annotated by \verb|@cps[Ans]|. 
This annotation denotes the final returned value of @ndcps@ is type @Ans@.
The @f@ introduced by the @shift@ operator is the delimited continuation from the call-site of @ndcps@.

@@

Now we have arrive the end of this series of transformations. Starting from an small-step abstract abstract machine, 
eventually we obtain an big-step definitional interpreter with pushdown control-flows and written in direct-style.

>>>>

To run the program, we first use the @inject@ function (below) to construct an initial machine
state given a closed expression @e@. The initial state contains an empty environment,
store, and stack.
\begin{lstlisting}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{lstlisting}

The @drive@ function is then used to evaluate
to a final state by iteratively applying @step@ on the current state until a state is reached
which the control is an atomic expression and the continuation stack is empty.
Naturally, we can then extract the value from the final state at last.

@@

In the definition of @State@, the continuation store disappears, and
component @konts@ becomes a list of frames. 
The other components remain unchanged as in the Section \ref{aam}.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentions that for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
This poses a problem, as the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (i.e., the @Frame@ in our program) is \emph{not}
nondeterministic, and the worklist @todo@ actually implicitly handles the nondeterminism.
Thus, if we simply refunctionalize the frames to higher-order functions, 
it does not help us to move toward abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all nondeterministic choices.
This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second a \emph{nondeterministic continuation}.

@@

Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the function's formal argument,
and the store, time, and frames at the fork point.
In the definition of @NDState@, an auxiliary function @toState@ is added that
converts itself to @State@.

@@

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should
determine the next state indicated by @NDCont@.
Thus, we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we have tried all the closures of this fork point and should move to the
next fork point. Since the @NDCont@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The nondeterministic continuation @new_ndk@ is also updated by the removal of that closure.
\end{itemize}

@@

With this, we have a single function to perform both abstract evaluation and the collection of
intermediate states. Given @inject(e)@ as an initial state and an empty set as the
initial set of states reached, we can easily define the entrance function @analyze@ as
follows:

@@

Function @continue@ is mainly what we have seen for the second case in
@drive_step@, except that @ndcontinue@ is called when
the normal continuation is empty.
Since we are analyzing a language in ANF, the first-order data representation of
continuations is simplified to a Scala @List@.
An empty list @Nil@ represents the halt of a computation path, whereas
a cons @::@ means the top frame of the list holds the binding variable and
evaluation context.
An analogy to this in standard CEK machines would be an abstract class (e.g., @cont@) with
three variants: @halt@, @arg(value, cont)@, and @arg(exp, env, cont)@.

@@

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ implements the nondeterministic
evaluation, given the defined @nd@, the evaluation function @aeval@ can be easily
defined. In the case that @e@ is a @let@ expression,
we invoke @nd@ twice and perform a nesting two-step, depth-first evaluation
over possible values of @App(f, ae)@.
The values of branches from a fork point will be accumulated, and then
returned to the upper fork point.

@@

Based on the refunctionalized AAM from the last section, we use the same nondeterministic
operator @nd@, but further extend it to a fixed-point cached evaluation in this section.
The fixed-point caching algorithm performs a sound over-approximation of all possible
concrete reachable paths and values, and also prevents non-termination when analyzing 
programs that may diverge.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantics artifact, then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order non-deterministic evaluation through higher-order functions representing continuations, 
but also drives us to use a new caching algorithm.

@@

\textbf{Nondeterminism Operators, Again.}
The first operator @nd@ is simplified from the previous one, but specialized with type @Ans@ and @Cache@. 
We may still keep the types generic, but the presentation clearly, exposing @Cache@ explictly
reveals that cache should be monotonically accumulated to the rest of computation.
Now @nd@ only takes a function @f@ that applies on type @(T, Cache)@ where the value of type @T@ comes from the set,
and @nd@ performs recursive call on itself over elements in the set @ts@.
We also move the accumulation operation (@++@) inside of @nd@.

@@

The function @ndcps@ takes two arguments: a set of elements with type @T@, 
and an initial accumulated value of type @Ans@.
The @ndcps@ iteratively returns an element from the set along with the latest cache, 
and the return type is also annotated by \verb|@cps[Ans]|. 
This annotation denotes the final returned value of @ndcps@ is type @Ans@.
The @f@ introduced by the @shift@ operator is the delimited continuation from the call-site of @ndcps@.

@@

Now we have arrived the end of this series of transformations. Starting from an small-step abstract abstract machine, 
eventually we obtain an big-step definitional interpreter with pushdown control-flows and written in direct-style.
